Tools for identifying unexpectedly low microfossil count sums

### 2023-01-01 manuscript/count_check_MS.qmd
tar_load

---
title: "Tools for identifying unexpectedly low microfossil count sums"
author:
  - name: Richard J. Telford
    email: richard.telford@uib.no
    affiliation: University of Bergen
    footnote: Corresponding Author
address:
  - code: University of Bergen
    address: Department of Biological Sciences, University of Bergen and Bjerknes Centre for Climate Research, Post Box 7803, N-5020 Bergen, Norway
abstract: |
  Microfossil counts are a key data type in palaeoecology. Recent work has raised the possibility that some authors might misreport an important quality control parameter, the counts sums, occasionally dramatically so, and perhaps other aspects of the counts. This paper explores the statistical properties of ecological and palaeoecological species counts and uses the findings to develop methods to flag count or percentage data with unexpected distributions. Several assemblage datasets that fail these tests, and potentially have misreported counts or count sums, are identified.
date: today
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{pdflscape}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}   
format: pdf
documentclass: article
classoption: a4paper
bibliography: extra/countMS2.bib
csl: extra/elsevier-harvard_rjt.csl
---

@@

# Packages loaded for convenience when knitting without running r_make
library("tidyverse")
library("rjt.misc")

@@

tar_load(bird_singleton_summary, bird_summ)
tar_load(testate_summ)
tar_load(last_chance_summ)
tar_load(pollen_summ, hershop_pollen_summ, wildhorse, pollen_summ1, pollen_summ2, pollen3, pollen1, pollen2)
tar_load(owens_summ1, owens_summ, owens_est_summ)
tar_load(diatom1_summ, diatom2_summ, chironomid1_summ, marine1_summ, chironomid2_summ)

@@

```{r}
#| label: fig-bird-recount
#| fig.cap: Effect of count sum on the proportion of bird counts from the North American
#|   breeding bird survey without singleton at the order level. Results are the mean
#|   of ten trials with 95% confidence interval.
#| fig.width: !expr 80/25.4
#| fig.height: !expr 70/25.4
tar_read(bird_recount_singleton_plot)
```

@@

Dataset chironomid1 has `r chironomid1_summ$nsamples` assemblages.
Although the associated paper reports that assemblages with count sums below 50 were discarded, the direct search method estimates that `r chironomid1_summ$n_below50 |> words` assemblages have count sums between `r chironomid1_summ$min_below50` and `r chironomid1_summ$max_below50`.
These assemblages are diverse (richness between `r chironomid1_summ$richness |> paste(collapse = " and ")` taxa), so it is unlikely that the true counts are double that estimated here and the greatest common divisor is two.

>>>>

---
title: "Tools for identifying unexpectedly low microfossil count sums"
author:
  - name: Richard J. Telford
    email: richard.telford@uib.no
    affiliation: University of Bergen
    footnote: Corresponding Author
address:
  - code: University of Bergen
    address: Department of Biological Sciences, University of Bergen and Bjerknes Centre for Climate Research, Post Box 7803, N-5020 Bergen, Norway
abstract: |
  Microfossil counts are a key data type in palaeoecology. Recent work has raised the possibility that some authors might misreport an important quality control parameter, the counts sums, occasionally dramatically so, and perhaps other aspects of the counts. This paper explores the statistical properties of ecological and palaeoecological species counts and uses the findings to develop methods to flag count or percentage data with unexpected distributions. Several assemblage datasets that fail these tests, and potentially have misreported counts or count sums, are identified.
date: today
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{pdflscape}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}   
format: pdf
documentclass: article
classoption: a4paper
bibliography: extra/countMS2.bib
csl: extra/elsevier-harvard_rjt.csl
execute: 
  echo: false
  error: false
---

@@

# Packages loaded for convenience when knitting without running r_make
library("tidyverse")
#library("rjt.misc")

@@

tar_load("summarised_bird_singletons")
tar_load("summarised_bird")
tar_load("testate_summ")
tar_load("last_chance_summ")
tar_load("pollen_summ")
tar_load("hershop_pollen_summ")
tar_load("wildhorse")
tar_load("pollen_summ1")
tar_load("pollen_summ2")
tar_load("pollen3")
tar_load("pollen1")
tar_load("pollen2")
tar_load("owens_summ1")
tar_load("owens_summ")
tar_load("owens_est_summ")
tar_load("diatom1_summ")
tar_load("diatom2_summ")
tar_load("chironomid1_summ")
tar_load("marine1_summ")
tar_load("chironomid2_summ")

@@

```{r}
#| label: fig-bird-recount
#| fig.cap: Effect of count sum on the proportion of bird counts from the North American
#|   breeding bird survey without singleton at the order level. Results are the mean
#|   of ten trials with 95% confidence interval.
#| fig.width: !expr 80/25.4
#| fig.height: !expr 70/25.4
tar_read(recounted_bird_singletons_plot)
```

@@

Dataset chironomid1 has `r chironomid1_summ$nsamples` assemblages.
Although the associated paper reports that assemblages with count sums below 50 were discarded, the direct search method estimates that `r chironomid1_summ$n_below50 |> words()` assemblages have count sums between `r chironomid1_summ$min_below50` and `r chironomid1_summ$max_below50`.
These assemblages are diverse (richness between `r chironomid1_summ$richness |> paste(collapse = " and ")` taxa), so it is unlikely that the true counts are double that estimated here and the greatest common divisor is two.
